% !TEX root = ../main.tex
%----------------------------------------------------------
% 00-abstract.tex — Abstract
%----------------------------------------------------------

\begin{abstract}
Communication of risk under high uncertainty, such as severe convective weather, air pollution, etc., often uses sharp probabilities. Possibility theory (the idea that probabilities themselves have uncertainty) offers a framework for measuring multifaceted aspects of hazard forecasting. A possibility framework allows a stepping-down to probabilistic or categorical/binary predictions where needed, hence an appropriate amount of complexity may be retained. In pursuit of model evaluation and optimisation, this paper presents a dedicated verification methodology. A \textit{subnormal} possibility
distribution $\pi$ over multiple outcomes---one whose maximum falls below unity---is framed as three forecast components: possibility $\poss$ (how
plausible is each outcome?), ignorance $\ign$ (how much of the
plausibility budget is unassigned?), and conditional necessity $\Nc$
(amongst covered outcomes, which dominates?).  Ignorance $\ign$ has
no probabilistic counterpart: probability, which must sum to unity,
cannot distinguish confident forecasts from uncertain ones.
This paper presents a five-number scorecard for possibilistic forecasts that evaluates
this forecast after observing the outcome, comprising
depth-of-truth ($\alpha^*$), diffuseness ($\eta$), support margin
($\delta$), conditional necessity of truth
($\Nc^*$), whilst also carrying $\ign$ as a neutral metric. A further possibility-to-probability conversion explicitly preserves
ignorance. This allows comparison of probabilistic forecasts from the same system using, e.g., Brier Score or (Shannon) information gain. Finally, the same dataset is reduced further to categories. These may be binary (no expression of uncertainty); herein, they follow Storm Prediction Center (SPC) convention of categories to demonstrate severe-weather prediction with a possibilistic system. Combined, metrics on this scorecard diagnose a range of failure modes captured only with insight from multiple metrics. Visualisations created from a synthetic dataset showcases benefits and interpretations of the framework, and aim to persuade the reader that ignorance is better expressed than repressed.   
\end{abstract}

\keywords{risk communication, forecast verification, possibility theory, uncertainty, scoring rules, severe weather}
