% !TEX root = ../main.tex
%----------------------------------------------------------
% 06-tripartite-value.tex — Section 5: Three-Component Value
% MIGRATED from preprint A5-verification-framework.tex §E.6
%    + A6-plain-language-verification.tex three-lanes section
% Removed: Clyfar-specific language, appendix framing
% Ozone categories → SPC categories
%----------------------------------------------------------

\section{Three-Component Value}\label{sec:tripartite_value}

% [PLACEHOLDER: Opening paragraph — a reviewer might ask "Why do you need
%  all three components?  Perhaps ignorance is redundant once you have
%  possibility and necessity."  This section lays out three hypothesis
%  tests, each isolating one component and asking whether it adds
%  information the other two do not provide.]

%----------------------------------------------------------
\subsection{Three Hypothesis Tests}\label{sec:three_tests}

The five-number scorecard (Section~\ref{sec:native_scorecard}) can be
used to demonstrate that each of the three named
components---possibility, ignorance, and conditional necessity---independently
contributes verification value.

\paragraph{Test 1: Possibility compatibility ($\alpha^*$, $\delta$).}

\textit{Question:} ``Does the system point at the truth better than
climatology or persistence?''

\textit{How to test:} Compare the forecasting system's resolution gap
($\delta = \alpha^* - \eta$) against the same score for climatological
and persistence baselines.  If the system's $\delta$ is consistently
higher, the possibility component has independent skill.

\textit{Met analogue:} Comparing a model's Brier skill score against
persistence---``Is your resolution better than a na\"ive forecast?''

% [PLACEHOLDER: Discuss expected behaviour for SPC outlooks — a system
%  that consistently assigns higher possibility to the observed category
%  than to the average category has positive resolution gap.]

\paragraph{Test 2: Ignorance self-awareness (\ign{}).}

\textit{Question:} ``When the system says it is uncertain, is it
actually performing worse?''

\textit{How to test:} Stratify forecasts by \ign{} (low vs.\ high
ignorance).  Check whether mean $\delta$ is lower in high-ignorance
cases.  If yes, the system's self-reported uncertainty tracks actual
forecast difficulty.

\textit{Met analogue:} The \textbf{spread--skill relationship} in
ensemble forecasting.  Stratify by ensemble spread and check whether
high-spread cases have higher error.  This test asks the same question
about \ign{}: is it an honest uncertainty signal, or just noise?

% [PLACEHOLDER: Discuss implications — if H_Pi does NOT track forecast
%  difficulty, the subnormality signal is uninformative and normalisation
%  is costless.  If it DOES track difficulty, normalisation destroys
%  operationally valuable information.]

\paragraph{Test 3: Conditional necessity reliability ($\Nc^*$).}

\textit{Question:} ``When the system says it is sure, is it actually
right more often?''

\textit{How to test:} When $\Nc(c) > \tau$ for some threshold $\tau$,
check whether the conditional hit rate for category $c$ exceeds the base
rate.  If yes, high-necessity forecasts identify genuinely confident
predictions---not just lucky ones.

\textit{Met analogue:} \textbf{Reliability conditional on confidence.}
Meteorologists check this when they ask ``when the ensemble gives
${>}80\%$ chance of severe weather, does it actually verify
${>}80\%$ of the time?''  This test asks: ``when the system says
$\Nc > \tau$, does the truth verify at a rate above climatology?''

% [PLACEHOLDER: Discuss sample-size requirements.  At pilot scales
%  (~90 station-days), these tests are supplemented by illustrative case
%  studies.  Recommend minimum sample sizes for each test.]

%----------------------------------------------------------
\subsection{Three Verification Lanes}\label{sec:three_lanes}

A single metric cannot capture full uncertainty quantification from a
subnormal possibility distribution.  Table~\ref{tab:three_lanes}
summarises the three verification lanes and their respective blind spots.

\begin{table}[htbp]
\centering
\small
\caption{Three verification lanes and their blind spots.}
\label{tab:three_lanes}
\begin{tabularx}{\columnwidth}{%
    >{\raggedright\arraybackslash}p{0.18\columnwidth}XX}
\toprule
Lane & Answers the question & What it misses \\
\midrule
\textbf{Deterministic} (RMSE, bias)
    & ``Was the point forecast close?''
    & All uncertainty information---a perfect point forecast says nothing
        about confidence. \\
\textbf{Probabilistic} (IG via bridge)
    & ``Was the probabilistic forecast informative?''
    & The bridge is lossy---it compresses the full possibility shape
        into $n{+}1$ probabilities.  Subnormality becomes a single
        ignorance bin. \\
\textbf{Native possibilistic} (scorecard)
    & ``Was the possibility distribution well-shaped, sharp, and
        honest?''
    & No connection to probabilistic skill---a well-shaped distribution
        might still give poor probabilities after conversion. \\
\bottomrule
\end{tabularx}
\end{table}

Together, the three lanes catch failures that any single lane would miss:

\begin{itemize}[leftmargin=2em]
\item A forecast with good deterministic scores but flat possibility
    (high $\eta$) and poor IG is a broken clock that happens to be
    right.
\item A forecast with great IG but high ignorance and low $\Nc^*$ got
    lucky via the bridge---the raw system was hedging.
\item A forecast with sharp possibility and high $\Nc^*$ but poor IG has
    a bridge problem---the conversion is distorting the signal.
\end{itemize}

% [PLACEHOLDER: 1–2 paragraphs expanding on how a forecaster or
%  verification analyst would use the three lanes in practice.  Suggest a
%  diagnostic workflow: start with the deterministic lane for quick
%  screening, move to the probabilistic lane for calibration assessment,
%  then examine the native possibilistic lane for deeper insight into
%  system behaviour.]
